Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations

### 2021-12-07 analysis/paper/paper.Rmd
updated KS address

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [biomark, mhe]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Under A Bridge, Seattle, Washington, 00000, USA
  - mhe: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

The Upper Salmon River major population group (MPG) supports eight independent, extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* each providing important ecological, social, and economic benefits to the region. The distinct populations include Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]; all of which have become depleted in recent decades. Reduction in overall population size, and more specifically, declines in survival of juvenile Chinook Salmon have been associated with hydrosystem passage [@Schaller2014], ocean conditions, regional climate [@Welch2021; @Gosselin2021], and downstream predation by native and non-native piscivorous fish [@Poe1991; @Friesen1999; @Harvey2004]. Anthropogenic changes to freshwater habitat including removal of beavers, river channel simplification, water withdrawals, and the proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*) among others [@IdahoOSCTeam2019] have also been attributed to the reductions in the population. Moreover, the abundance of returning adults is futher impacted by ocean and mainstem harvests, poor ocean conditions, and changes to the spawning migration corridor [@FCRPSBiOp2020]. Each of these factors have contributed, to varying extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Upper Salmon Chinook Salmon abundance from the myriad of human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in rehabilitation of tributary ecosystems as well as investing in the study and evaluation of many factors attributed to the species decline [@Roni].

One potentially important, but perhaps under-appreciated source of mortality for Upper Salmon Chinook Salmon is predation on emigrating juveniles by piscine fishes. Research in the lower mainstem Snake and Columbia rivers has estimated that during peak emigration, predation on out-migrating salmonids has a significant negative impact (most notably in the presence on non-native piscivores) on the overall population and success of population recovery [@Fresh2003; @Winther2019]. However, these studies have focused on larger, mainstem and reservoir habitats and direct predation, but very little is known about the interaction between upstream, river dwelling piscine fishes, and their impact on salmonid recovery above reservoir systems of the Columbia and Snake Rivers [@Rubenson2020]. Adding to this, there is little consensus within the literature related to the magnitude of predation on salmonids in less altered, free flowing systems, with some suggesting lower levels of predation [@Brown1981; @Buchana1981] and others suggesting increased levels of predation can occur [@Tabor1993; @Ward1995; @Shively1996; @Zimmerman1999].

>>>>

---
title: "Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations"
author:
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: true
  - Nicholas J. Porter:
      email: Nicholas.Porter@merck.com
      institute: [biomark]
      correspondence: false
  - Kevin E. See:
      email: Kevin.See@dfw.wa.gov
      institute: [biomark, wdfw]
      correspondence: false
  - Michael S. Hall:
      email: Mike.Hall1@merck.com
      institute: [biomark]
      correspondence: false
  - Tara E. Blackman:
      email: tara.blackman@mthoodenvironmental.com
      institute: [mhe_sandy]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@mthoodenvironmental.com
      institute: [biomark, mhe_mccall]
      correspondence: false
institute:
  - biomark: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - wdfw: Washington Department of Fish and Wildife, Fish Program, Science Division, 1111 Washington Street NE, Olympia, Washington, 98501, USA
  - mhe_mccall: Mount Hood Environmental, PO Box 4282, McCall, Idaho, 83638, USA
  - mhe_sandy: Mount Hood Environmental, 39085 Pioneer Boulevard \#100 Mezzanine, Sandy, Oregon, 97055, USA 
date: "`r format(Sys.time(), '%d %B, %Y')`"
always_allow_html: true
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua      
fontsize: 12pt
mainfont: Times New Roman
bibliography: 
- references.bib      # Kevin's library
- AckermanLibrary.bib # Mike's library
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Abstract text here. 
keywords: |
  northern pikeminnow; Chinook salmon; predation; mark-recapture; bioenergetics
highlights: |
  These are the highlights. 
---

@@

The Upper Salmon River major population group (MPG) supports eight independent, extant spring/summer Chinook Salmon *Oncorhynchus tshawytscha* each providing important ecological, social, and economic benefits to the region. The distinct populations include Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017], all of which have become depleted in recent decades. Reduction in overall population size, and more specifically, declines in survival of juvenile Chinook Salmon have been associated with hydrosystem passage [@Schaller2014], ocean conditions, regional climate [@Welch2021; @Gosselin2021], and downstream predation by native and non-native piscivorous fish [@Poe1991; @Friesen1999; @Harvey2004]. Anthropogenic changes to freshwater habitat including removal of beavers, river channel simplification, water withdrawals, and the proliferation of non-native species (e.g., non-native coastal rainbow trout *O. mykiss irideus* and brook trout *Salvelinus fontinalis*) among others [@IdahoOSCTeam2019] have also been attributed to the reductions in the population. Moreover, the abundance of returning adults is futher impacted by ocean and mainstem harvests, poor ocean conditions, and changes to the spawning migration corridor [@FCRPSBiOp2020]. Each of these factors have contributed, to varying extents, to reduced adult escapement, the primary metric used to assess population viability. In response to the decline in Upper Salmon Chinook Salmon abundance from the myriad of human activities and associated habitat degradation, action agencies have attempted to improve juvenile survival and adult spawning conditions by investing in rehabilitation of tributary ecosystems as well as investing in the study and evaluation of many factors attributed to the species decline [@Roni2018].

One potentially important, but perhaps under-appreciated source of mortality for Upper Salmon Chinook Salmon is predation on emigrating juveniles by piscine fishes. Research in the lower mainstem Snake and Columbia rivers has estimated that during peak emigration, predation on out-migrating salmonids has a significant negative impact (most notably in the presence on non-native piscivores) on the overall population and success of population recovery [@Fresh2003; @Winther2019]. However, these studies have focused on larger, mainstem and reservoir habitats and direct predation, but very little is known about the interaction between upstream, river dwelling piscine fishes, and their impact on salmonid recovery above reservoir systems of the Columbia and Snake Rivers [@Rubenson2020]. Adding to this, there is little consensus within the literature related to the magnitude of predation on salmonids in less altered, free flowing systems, with some suggesting lower levels of predation [@Brown1981; @Buchanan1981] and others suggesting increased levels of predation can occur [@Tabor1993; @Ward1995; @Shively1996; @Zimmerman1999].
